---
title: "Automation Processing"
author: "Laura Palacios"
date: '2022-11-13'
output: html_document
---

```{r setup, include=FALSE}
# Libraries
library(dataRetrieval)
library(janitor)
library(knitr)
library(mapview)
library(sf)
library(tidyverse)
library(xgboost)

# Not currently on CRAN - devtools::install_github("scoyoc/climateAnalyzeR")
library(climateAnalyzeR)

opts_chunk$set(echo = TRUE)
```

## E. coli 235 Model {.tabset}

### Model Processing

I will need to read in:

1. PreviousTmin(Climateanalyzer - lagged raw value)
2. Discharge_CFS (USGS WaterData - Raw value)
3. Stage (USGS Waterdata - Calculated Categorical
4. NinXTS (NOAA?)
5. TOD (Use Current?)
6. DistFromSonoita (Think about this one. Can I calc using latlong?) https://stackoverflow.com/questions/32363998/function-to-calculate-geospatial-distance-between-two-points-lat-long-using-r

Default values for data pulls - only need to do this once

```{r include=FALSE}

# Find start and end dates to use for data pulls
sysDate1 <- Sys.time()

```

#### 1 - Previous T min (2 days prior). I need to lag this one more time in the processing due to the outputs of climate analyzer. This was previously used as the 30 day precipitation values, but the rain gage data is not trustworthy anymore due to birds and mowing vehicles.

```{r}
# Pulls in current and 1 previous year's data (to account for January calcs)
get.Tmin <- function(sysDate) {
  formattedEndYear <- as.numeric(format(sysDate, "%Y"))
  TMin <- climateAnalyzeR::import_data("daily_wx"
                                       , station_id = 'KA7WSB-1'
                                       , start_year = formattedEndYear-1
                                       , end_year = formattedEndYear
                                       , station_type = 'RAWS')
  
  Var_TMin <- as.numeric(unlist(TMin %>%
    mutate(DateasDate = as.POSIXct(TMin$date, format = "%m/%d/%Y")) %>%
    subset(DateasDate == as.Date(sysDate) - 2) %>%
    select(tmin_f)))
}

Var_TMin <- get.Tmin(sysDate1)

```

#### 2 - Discharge data

Use dataRetrieval to pull in the data from the USGS waterdata website.
```{r}
get.DischargeCFS <- function(sysDate) {
  startDate <- as.Date(format(sysDate1,'%Y-%m-%d')) - 31
  endDate <- as.Date(format(sysDate1,'%Y-%m-%d')) - 1
  USGSRaw <- readNWISuv(siteNumbers = '09481740', c('00060','00065'), startDate,endDate, tz = 'America/Phoenix')
  
  tail(USGSRaw$X_00060_00000, n=1)
}

Var_Discharge_CFS <- get.DischargeCFS(sysDate1)
```

#### 3 - Stage
```{r}
get.stage <- function(sysDate) {
  startDate <- as.Date(format(sysDate1,'%Y-%m-%d')) - 31
  endDate <- as.Date(format(sysDate1,'%Y-%m-%d')) - 1
  USGSRaw <- readNWISuv(siteNumbers = '09481740', c('00060','00065'), startDate,endDate, tz = 'America/Phoenix')

  # Create quantiles for categorization
  CFS_Quantiles<- quantile(USGSRaw$X_00060_00000, na.rm = TRUE)

  # Determine the difference between prior reading and current.
  USGSRaw <- USGSRaw %>% 
    mutate(DisDif = X_00060_00000 - lag(X_00060_00000))

  # This will create a binary variable or either rise of fall. Rise = 1, fall = 0. It will allow me to more easily create summary statistics.
  USGSRaw$DisDif2 <- ifelse(USGSRaw$DisDif>0,1,0)

  # Create a numeric classifier. 
  # 1 = Low Flow, 2 = Base flow, 3 = High and Rising Flow 4 = High and Falling Flow
  USGSRaw$Stage <- ifelse(USGSRaw$X_00060_00000 <=CFS_Quantiles[2], 1, 
    ifelse(USGSRaw$X_00060_00000 > CFS_Quantiles[2] & USGSRaw$X_00060_00000 <= CFS_Quantiles[4],2,
    ifelse(USGSRaw$X_00060_00000 > CFS_Quantiles[4] & USGSRaw$DisDif2 == 1,3,
    ifelse(USGSRaw$X_00060_00000 > CFS_Quantiles[4] & USGSRaw$DisDif2 == 0,4, NA))))


# Create the stage variable.
tail(USGSRaw$Stage, n=1)
}

Var_Stage <- get.stage(sysDate1)
```

#### 4 - NinXTS
```{r}
get.NinXTS <- function(sysDate) {
  formattedEndYear <- as.numeric(format(sysDate, "%Y"))
  formattedMonth <- as.numeric(format(sysDate,"%m"))
  
# Bring in the website data
url <- "https://origin.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/ONI_v5.php"

NinXTS <- url %>%
  rvest::read_html()

# Grab the data table
NinText <- rvest::html_table(rvest::html_nodes(NinXTS, xpath = './/table[4]//table[2]'))

# Convert ONI index to dataframe
NinTable <- as.data.frame(NinText[1]) %>%
  row_to_names(row_number = 1) %>%
  mutate(Year = as.numeric(Year)) %>%
  drop_na(Year)

# I need to do either month-of or last non-na value to account for delays. 
formattedMonth <-12

NinVal <- NinTable %>%
  subset(2022 == Year) %>%
  select(case_when(formattedMonth == 1 ~ "NDJ",
                   formattedMonth == 2 ~ "DJF",
                   formattedMonth == 3 ~ "JFM",
                   formattedMonth == 4 ~ "FMA",
                   formattedMonth == 5 ~ "MAM",
                   formattedMonth == 6 ~ "AMJ",
                   formattedMonth == 7 ~ "MJJ",
                   formattedMonth == 8 ~ "JJA",
                   formattedMonth == 9 ~ "JAS",
                   formattedMonth == 10 ~ "ASO",
                   formattedMonth == 11 ~ "SON",
                   formattedMonth == 12 ~ "OND")) %>%
  unlist()

PrevVal <- NinTable %>%
  subset(2022 == Year) %>%
  select(case_when(formattedMonth == 2 ~ "NDJ",
                   formattedMonth == 3 ~ "DJF",
                   formattedMonth == 4 ~ "JFM",
                   formattedMonth == 5 ~ "FMA",
                   formattedMonth == 6 ~ "MAM",
                   formattedMonth == 7 ~ "AMJ",
                   formattedMonth == 8 ~ "MJJ",
                   formattedMonth == 9 ~ "JJA",
                   formattedMonth == 10 ~ "JAS",
                   formattedMonth == 11 ~ "ASO",
                   formattedMonth == 12 ~ "SON",
                   formattedMonth == 1 ~ "OND")) %>%
  unlist()

if_else(!is.na(NinVal), NinVal, PrevVal)                 
}

Var_NinXTS <- get.NinXTS(sysDate1)

```


#### 5 - TOD
```{r}
#modified from https://stackoverflow.com/questions/49370387/convert-time-object-to-categorical-morning-afternoon-evening-night-variable

get.TOD <- function(sysTime) {

# Create categorical variables
currenttime <- as.POSIXct(sysTime, format = "%H:%M") %>% format("%H:%M:%S")

currenttime <- cut(chron::times(currenttime) , breaks = (1/24) * c(0,5,11,16,19,24))
Var_TOD <- c(4, 1, 2, 3, 4)[as.numeric(currenttime)]
}


Var_TOD <- get.TOD(sysDate1)
```

#### 6 - Distance from Sonoita - Roughly estimated linear distance from Sonoita input.

Create the matrix to put through the model.
```{r}
spatiallocs <- read_sf("Data/Processed/Attributed_Location/ecoli_UniqueLocs.shp")

# Retrieve all variables using the functions
sysDate1 <- Sys.time()



predictionDF <- as.data.frame(spatiallocs)
predictionDF$PreviousTmin <- c(Var_TMin)
predictionDF$Discharge_CFS	<- c(Var_Discharge_CFS)
predictionDF$Stage	<- c(Var_Stage)
predictionDF$NinXTS	<- c(Var_NinXTS)
predictionDF$TOD <- c(Var_TOD)

predictionDF <- predictionDF %>%
  rename(DistFromSonoita = DistCatego) %>%
  select(PreviousTmin, Discharge_CFS, Stage, NinXTS, TOD, DistFromSonoita)

```

This is pulled from the raw data. I should be able to attribute it with the rest of the information, run the model, and create a prediction for each spot. Maybe use green/red as the coloring?

```{r}
XGBModel <- xgb.load('Data/Processed/XGBmodel')
predictionDM <- data.matrix(predictionDF)
pred <- predict(XGBModel,predictioDM)
pred <-  as.numeric(pred > 0.4)
spatiallocs$pred2 <- c(pred)
spatiallocs$pred <- ifelse(spatiallocs$pred2 > 0, "Bacteria Level >235  Likely", "High Bacteria levels > 235 not predicted")
spatiallocs <- spatiallocs %>%
  select(-DistCatego) %>%
  rename(Samplers = Organiza_2, Prediction = pred)
```


```{r}
#XGB575Model <- xgb.load("Data/Processed/final575model")


```



### Plots
```{r}
mapview(spatiallocs, z = "Prediction", col.regions=list("red","blue"),col=list("red","blue"),layer.name = "Predictions")
```





```{r}
library(leaflet)
library(sp)

ColorPal <- colorFactor(palette = c('Green', 'Red'), levels = c(0,1))

m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addCircleMarkers(
    data = spatiallocs
    , popup = paste0( "Sampling Organization(s):"
                 ,spatiallocs$Samplers
                 , "<br>"
                 ,spatiallocs$Prediction)
    , color = ColorPal(spatiallocs$pred2)
               )
m  # Print the map
```





